
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      
  warning = FALSE, 
  message = FALSE   
)
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
```


```{r}
flu_temp <- read_csv("/Users/naomioluyemi/Downloads/PredictUS_Tx/Clean\ data/predictus_tx_flu_temp.csv")
 

# checking NAs
colSums(is.na(flu_temp))

flu_clean <- flu_temp %>%
  filter(!is.na(flu_percent) & !is.na(avg_temp))
nrow(flu_clean)

```

```{r}
# Providing the model with a one and two-week memory of prior surveillance data
flu_clean <- flu_clean %>%
  arrange(date) %>%
  mutate(
    lag1_flu = lag(flu_percent, 1),
    lag2_flu = lag(flu_percent, 2),
    lag1_temp = lag(avg_temp, 1)
  ) %>%
  filter(!is.na(lag2_flu))

nrow(flu_clean)

# verifying lag feature
flu_clean %>%
  select(date, flu_percent, lag1_flu, lag2_flu, avg_temp, lag1_temp) %>%
  head(10)
```

```{r}
# Scatter plot showing relationship between temperature and flu ER visits
ggplot(flu_clean, aes(x = avg_temp, y = flu_percent)) +
  geom_point(alpha = 0.4, color = "steelblue") + 
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Temperature vs Flu Percent in Texas (2015 -2025)",
    x = "Average Weekly Temperature (°F)",
    y = "Flu Percent of ER Visits"
  ) +
  theme_minimal()
```

Red line shows the trend (as temperature rises, flu visits decrease) while the gray shading shows the uncertainty range around the trend line.

```{r}
# Flu percent over time to show seasonal trends
ggplot(flu_clean, aes(x = date, y =flu_percent)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Weekly Flu Percent in Texas ER Visits (2015 - 2025)",
     x = "Date",
    y = "Flu Percent of ER Visits"
  ) +
  theme_minimal()

```

Flu percent rises every winter and drops every summer. This confirms seasonal pattern which justifies using temperature as a predictor. There is a notable spike in 2020 which likely reflects COVID-19 overlap in ER reporting.

```{r}
# Temperature over time to show seasonal cycle
ggplot(flu_clean, aes(x = date, y = avg_temp)) +
  geom_line(color = "tomato") +
  labs(
    title = "Weekly Average Temperature in Texas (2015 - 2025)",
    x = "Date",
    y = "Average Temperature (°F)"
  )
theme_minimal()

```
Temperature follows a clear seasonal cycle in Texas. When temperature hits winter lows, flu percent is at its highest. This inverse relationship visually supports our core hypothesis.


```{r}
# Linear Regression Model
model <- lm(flu_percent ~ avg_temp + lag1_flu +lag2_flu + lag1_temp,
            data = flu_clean)
summary(model)

```
#### Interpretation
R-squared: 0.9267; the model explains 92.7% of the variation in flu ER visits.
avg_temp: -0.027904 (p = 0.00146). For every 1°F increase in temperature, flu percent drops by 0.028% (statistically significant)
lag1_flu: 1.202481 (p < 2e-16). Last week's flu activity is the strongest predictor. If flu was high last week, it will likely be high this week. 
lag2_flu: -0.292514 (p < 2e-16). Two weeks ago flu activity has a corrective effect; it slightly pulls the prediction back down but remains extremely significant.
lag1_temp: 0.016399 (p = 0.06356). Last week's temperature is borderline; it misses the 0.05 significance threshold.


### Threshold
```{r}
# Mean and sdv of flu_percent from training data
mu <- mean(flu_clean$flu_percent)
sdv <- sd(flu_clean$flu_percent)

# finding optimal surge thresholds
k_values <- seq(1.0, 2.5, by = 0.1)

# results table
# F1 score balances two concerns:
# Sensitivity: are we catching real surges?
# Specificity:are we avoiding false alarms? 

results <- data.frame(k = k_values, sensitivity = NA, specificity = NA, f1 = NA)

for (i in seq_along(k_values)) {
  threshold <- mu + k_values[i] * sdv
  
  predicted_surge <- as.numeric(fitted(model) >= threshold)
  actual_surge <- as.numeric(flu_clean$flu_percent >= threshold)
  
  tp <- sum(predicted_surge == 1 & actual_surge == 1)
  fp <- sum(predicted_surge == 1 & actual_surge == 0)
  fn <- sum(predicted_surge == 0 & actual_surge == 1)
  tn <- sum(predicted_surge == 0 & actual_surge == 0)
  
  sens <- ifelse((tp +fn) > 0, tp / (tp + fn), 0)
  spec <- ifelse((tn + fp) > 0, tn / (tn + fp), 0)
  f1 <- ifelse((2**tp + fp + fn) > 0, (2*tp)/(2*tp + fp +fn), 0)
    
  results$sensitivity[i] <- round(sens, 3)
  results$specificity[i] <- round(spec, 3)
  results$f1[i] <- round(f1, 3)
  
}


print(results)

```
Sensitivity 86.1%: the model catches 86 out of every 100 real surges
Specificity 99.6%: the model correctly identifies normal weeks 99.6% of the time
F1 = 0.899 which is a strong overall balance between both
Optimal k = 1.9


```{r}
# Youden Index
results$youden <- results$sensitivity + results$specificity - 1
print(results)
```

#### Surge Detection
```{r}
flu_clean <- flu_clean %>%
  mutate(surge_flag = as.numeric(flu_percent >= (mu + 1.9 * sdv)))

ggplot(flu_clean, aes(x = date, y = flu_percent)) +
  geom_line(color = "steelblue") +
  geom_point(data = filter(flu_clean, surge_flag == 1),
             aes(x = date, y = flu_percent),
             color = "red", size = 2) +
  geom_hline(yintercept = mu + 1.9 + sdv,
             linetype = "dashed", color = "red") +
  labs(
    title = "Flu Surge Detection in Texas (k = 1.9)",
    x = "Date",
    y = "Flu Percent of ER Visits",
    caption = "Red dots = surge weeks | Dashed line = surge threshold"
  ) +
  theme_minimal()

  
  
```


